Anxiety, avoidance, and sequential evaluation

### 2019-06-30 manuscripts/manuscript/manuscript.tex
intro completed. almost done discussion.

Though the anxiety disorders differ in their particular symptomology, and in the content and situations which elicit anxious responding, they all are characterized by aberrations in the processing of and response to threat\citep{dsm5}. In particular, at least three symptoms seem to manifest across anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted\citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, situations, and events\citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which indirectly works to maintain pathology by preventing the experience of non-outcomes of expected bad events\citep{Arnaudova2017}. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

These symptoms are particularly puzzling from a decision theoretic perspective\citep{huys2015}. Fear and avoidance of situations far antecedent of threat violates the basic logic of evaluation over sequential trajectories of action. Put another way, future threat need not necessarily influence decision making in the present. This is because avoidance is by nature protective: the ability to avoid in future means an agent need not also do so now. For instance, cars endanger pedestrians but can be reliably avoided by following traffic signals; given that, staying indoors offers little or no additional protection from accidents. This is an instance of a general property of evaluating actions in sequential decision making: such evaluation turns fundamentally on assumptions about subsequent events, including the agent's own subsequent choices. Typically, it is appropriate to assume that an agent will also make reward-maximizing (and harm-minimizing) choices down the line.

This line of reasoning hints that a fundamental deficit in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of threat and excessive avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. Trait and clinically anxious individuals rate the likelihood of future threatening scenarios higher than do non-anxious individuals\citep{???}. Most importantly, anxious individuals are more likely to believe that they are unable to handle and mitigate future threat\citep{???}.

Here, we develop this idea into a model of evaluation under pessimistic assumptions about future choices. We show that a single, localized deviation can explain a surprising range of features of anxious behavior including exaggerated threat appraisal, fear generalization, and persistent avoidance. This account also offers a new formalization of classic insights from the psychiatric literature about the central role of beliefs about control and self-efficacy in anxiety. Specifically, we demonstrate a model which distinctly lacks a belief in the efficacy of future action gives rise to the aforementioned symptoms. Our perspective is importantly different from previous computational accounts of beliefs about control in mood disorders (e.g. \cite{HuysDayan2009}), which neglected the sequential aspects of choice.

@@

An agent could expect to encounter dangerous future events under different assumptions about the future: either if the agent expects to fail to take the correct avoidance actions (i.e., expects to use a nonoptimal $\pi(a \mid s)$), or if the world's dynamics do not guarantee reliable avoidance even so (e.g., under stochastic or adversarial transition dynamics $P(s' \mid s,a)$). Consider an agent who has such pessimistic expectations about such events at future steps. Computing returns with such pessimistic predictions can help to quantify uncertainty in outcomes (i.e. to learn different points in the distribution of possible returns) \cite{deepmindstuff}, and choosing on this basis can be adaptive to ensure robustness such as under adversarial scenarios \cite{Garcia2015}.

@@

Central to anxiety disorders are symptoms including pessimistic inference, threat generalization, and excessive avoidance. We have presented a simple computational account suggesting how a single underlying misbelief can give rise to a range of these aberrations in learning and choice. Specifically, we showed how a failure to believe in the reliability of future self-action can effectively backpropagate negative value across states of the environment. This process results in a range of inferences and behaviors resembling those observed in clinical anxiety. Though this is by no means a complete account of anxiety, it ties together a surprisingly wide range of its symptoms.

One important ambiguity is that we have formalized pessimistic assumptions in terms of beliefs about the agent's own future actions: expecting failure, for instance, to avoid in future. Since Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$, pessimism can alternatively be encoded in this distribution (e.g., a belief that the world's response to one's choices is unpredictable or adversarial). Because the Bellman equation averages out both distributions, and because an unpredictable environment also reduces the efficacy of avoidance, either formulation can produce ultimately similar results. However from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs, which may be relevant, for instance, in guiding treatment.

Indeed, our account formalizes a longstanding strand of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, learned helplessness theory claims that pathological anxiety results from a belief that the environment is uncontrollable, such that threat cannot be effectively mitigated \cite{alloy1990}. In contrast, Bandura's self-efficacy theory of anxiety posited that a lack of confidence in one's own ability to effectively handle threat was instead central \cite{bandura1977}. As we note above, the present model (though stated here in terms echoing Bandura's) can accommodate either variant, and they are by no means mutually exclusive.

%% Can probably flesh this out. Cite problem orientation literature as evidence.

The two accounts described above are by no means the only possible reinforcement learning formalisms that can give rise to in threat-avoidance behavior. Traditionally, reinforcement learning algorithms aim to find reward-maximizing policies with respect to the expected (average) returns. However, in environments with the potential for catastrophic loss it may be preferable to optimize a policy with respect to an alternative objective function insofar that the expected return does not necessarily avoid rare yet large negative outcomes. Such risk-sensitive reinforcement learning algorithms have been the subject of much recent research \cite{geibel2005, morimura2012} and show threat-avoidance behaviors not unlike those presented here. An often reported clinical observation is that anxious individuals disproportionately dwell upon low chance yet high cost possibilities. \cite{Miceli2005} One speculative hypothesis is that anxious individuals optimize with respect to a more conservative risk criterion.

A related possibility concerns the notion of state itself. Here we assumed that states were fully observable, such that there was no uncertainty regarding the state an agent was currently occupying. In real-world environments, however, the state must be inferred from external and internal variables. As has been proposed elsewhere \cite{Paulus2012}, one possibility is that anxiety biases state inference resulting in erroneous beliefs about likely future events. There is evidence that anxious individuals are biased in their sampling and interpretation of incoming threat information \cite{Paulus2012}, suggesting that the avoidance behaviors may be a function of pessimistic beliefs about the current state of the environment.

% Citing the Wu stuff may be helpful here

Of course, none of the above constitutes a claim about the biological mechanisms underlying pessimistic sequential evaluation. One possibility is that these symptoms reflect biased model-based planning. Pathological worry is strongly associated with anxiety disorders and may be viewed as a biased form of future action planning. In line with the present results, higher levels of worry is associated with reduced self-efficacy beliefs, greater endorsement of statements concerning personal inadequacies in the face of threat, and that the association between perceived competence and worry is mediated by exaggerated threat appraisal (for review, see \cite{Berenbaum2010}). As such, the behaviors observed above may reflect biased sampling of possible future outcomes as a function of pessimistic beliefs about self-efficacy.

As alluded to above, a second possibility is that biased sequential evaluation may arise as a function of biased offline replay of events. During periods of rest and sleep, the hippocampus replays recent experiences. Recent work has shown that episodes are prioritized for replay based according to their likelihood in changing behavior, and these replay events iteratively backpropagate value to antecedent states. Biases in the tagging of memories for replay, as a function of either biased expectations of future state occupancy (need) or biased estimates in the value to be learned, may yield the propagation of threat value similarly as has been observed above.

It is important to note that we do not believe that this model of anxiety is equally explanatory across the family of anxiety disorders. Indeed, those disorders more strongly associated with disordered prospective cognition, such as generalized anxiety disorder and social anxiety disorder, are more likely to be better described by biased sequential evaluation. This account will have trouble explaining other features of anxiety disorders, such as the compulsive behaviors of obsessive-compulsive disorder and the memory disturbancse associated with post-traumatic stress disorder. Importantly, other psychiatric disorders with avoidance behaviors, such as eating disorders, may also potentially be well described by this model. As such, this account is transdiagnostic.

% this is probably fine?

Finally, we are not the first to propose a formal theory of control in psychiatry using MDPs. Huys\cite{HuysDayan2009} provided a computational account of learned helplessness through simple models of one-step action-outcome contingencies. Our accounts differ particularly in our exclusive focus on control in the sequential setting, which Huys did not address. Indeed, we propose that ultimately key to anxiety is precisely the way in which evaluation in sequential tasks is necessarily reliant on expectations about future events.

>>>>

Though anxiety disorders differ in their particular symptomology, and in the content and situations which elicit anxious responding, they all are similarly characterized by aberrations in the processing of and response to threat\citep{dsm5}. In particular, at least three symptoms are manifest across the anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted\citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, events, and thoughts\citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which often occurs well in advance of the materialization of actual threat\citep{Arnaudova2017}. Avoidance behaviors in anxiety are especially pernicious as they indirectly maintain anxiety by preventing learning from the non-occurrence of perceived threats. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

These symptoms are particularly puzzling from a decision theoretic perspective\citep{huys2015}. Fear and avoidance of situations far in the future violates the basic logic of evaluation over sequential trajectories of action. Put another way, distant threat need not necessarily impinge on decision making in the present. This is because avoidance is by nature protective: the ability to successfully avoid danger in the future means an agent need not also do so now. For instance, cars endanger pedestrians but can be reliably avoided by following traffic signals; given that, staying indoors offers little or no additional protection from accidents. This is an instance of a general property of evaluation in sequential decision making: the value of present action turns fundamentally on assumptions about subsequent events, which include the agent's own subsequent choices. Typically, it is appropriate to assume that an agent will continue make good (i.e. reward-maximizing/harm-minimizing) choices down the line.

This line of reasoning hints that a fundamental deficit in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of threat and excessive avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. Trait and pathologically anxious individuals judge future threat as more likely than do non-anxious individuals\citep{ButlerMathews1983, ButlerMathews1987, MacleodByrne1996}. Importantly, the development and maintenance of clinical anxiety is strongly tied to diminished perceived control\citep{bandura1977, barlow2002, gallagher2014a}, such that anxious individuals are more likely to endorse the belief that they are unlikely or unable to mitigate future threat. Indeed, a belief (or lack thereof) of one's ability to successfully navigate future danger is associated with anxiety\citep{davey1996, dugas1997}, and an increased belief in perceived control over threat is correlated with symptom reduction across the family of anxiety disorders\citep{gallagher2014b}.

Here, we develop this idea into a model of evaluation under pessimistic assumptions about future choices. We show that a single, localized deviation can explain a surprising range of features of anxious behavior including exaggerated threat appraisal, fear generalization, and persistent avoidance. This account also offers a new formalization of classic insights from the psychiatric literature about the central role of beliefs about control and self-efficacy in anxiety \citep{bandura1977, barlow2002}. Specifically, we demonstrate a model with a misbelief in the reliability of future self-action gives rise to the anxious symptoms. Our perspective is importantly different from previous computational accounts of beliefs about control in mood disorders (e.g. \cite{HuysDayan2009}), which neglected the sequential aspects of choice.

@@

An agent could expect to encounter dangerous future events under different assumptions about the future: either if the agent expects to fail to take the correct avoidance actions (i.e., expects to use a nonoptimal $\pi(a \mid s)$), or if the world's dynamics do not guarantee reliable avoidance even so (e.g., under stochastic or adversarial transition dynamics $P(s' \mid s,a)$). Consider an agent who has such pessimistic expectations about such events at future steps. Computing returns with such pessimistic predictions can help to quantify uncertainty in outcomes (i.e. to learn different points in the distribution of possible returns)\cite{bellemare2017}, and choosing on this basis can be adaptive to ensure robustness such as under adversarial scenarios\cite{Garcia2015}.

@@

Central to anxiety disorders are symptoms including exaggerated threat appraisal, threat generalization, and excessive avoidance. We have presented a simple computational account suggesting how a single underlying misbelief can give rise to these aberrations in learning and choice. Specifically, we showed how a failure to believe in the reliability of future self-action can effectively backpropagate negative value across states of the environment. This process results in a range of inferences and behaviors resembling those observed in clinical anxiety. Though this is by no means a complete account of anxiety, it ties together a surprisingly wide range of its symptoms.

One important ambiguity is that we have formalized pessimistic assumptions in terms of beliefs about the agent's own future actions: expecting failure in handling or avoiding future threat. Since Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$, pessimism can alternatively be encoded in this distribution (e.g., a belief that the world's response to one's choices is unpredictable or adversarial). Because the Bellman equation averages out both distributions, and because an unpredictable environment also reduces the efficacy of avoidance, either formulation can produce ultimately similar results. However from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs, which may be relevant, for instance, in guiding treatment.

Indeed, our account formalizes a longstanding strand of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, self-efficacy theory\citep{bandura1977} and the triple vulnerabilities model\citep{barlow2002} both posit that a reduced belief in the ability to effectively respond to future threat is involved in the genesis and maintenance of clinical anxiety. Similarly, the learned helplessness theories of anxiety\cite{alloy1990} claims pathological anxiety results from an uncertain belief in the controllability of the environment, such that future threat cannot be effectively mitigated or avoided. As we note above, the present model (though stated here in terms of self-efficacy) can accommodate either variant, and they are by no means mutually exclusive.

Distortions in beliefs about future action or transition probabilities are in no way the only means through which anxious behaviors may arise in a decision theoretic account. Traditionally, the goal of reinforcement learning algorithms is to a find reward-maximizing policy with respect to the expectation (average) of returns. However, in environments where the distribution of returns has high variance, such that there is potential for catastrophic loss, it may be instead preferable to learn a policy with respect to an alternative value function. The same is true when there is uncertainty in the transition probabilities, such that an agent may inadvertently step towards undesirable and/or dangerous segments of the environment. How to learn risk-sensitive and robust policies is the focus of much recent research \cite{morimura2012, chow2015, bellemare2017}; interestingly, these studies have reported agent behaviors not unlike those presented here. It is important to note that anxiety is correlated with disproportionate worry about low chance but high cost events\cite{Miceli2005}. One speculative hypothesis is that anxious individuals learn and evaluate action with respect  to a more pessimistic point in the distribution of expected returns.

A separate possibility concerns the notion of state. In the present work, we considered only environments with fully observable states; in other words, there was no uncertainty regard the state an agent was currently occupying. In the real-world, however, the state must be inferred from external and internal observations. One possibility is that anxiety is a function of biases in state inference\cite{Paulus2012}. For example, if an agent erroneously comes to the conclusion that she is more proximal to threat than is actually the case, then symptoms like exaggerated threat appraisal and avoidance should arise. Indeed, a bias in value functions could result in the diffuse spread of threat across objects\citep{norbury2018} and location\citep{schulz2018}, thereby causing otherwise benign states to be perceived as threatening.

None of the previous discussion constitutes a claim about the algorithmic and neural bases of pessimistic sequential evaluation. One possibility is that these symptoms reflect aberrations in model-based planning. Chronic worry is a symptom frequently observed in pathological anxiety, and could be construed as a deviant form of future action planning wherein the roll-out of future actions is biased towards bad- or worst-case contingencies. In line with the present results, higher levels of worry is correlated with reduced perceived control, increased doubt in self-efficacy in response to threat, and exaggerated threat appraisal\cite{Berenbaum2010}. In sum, anxious symptoms may in part result from biased planning where cost-minimizing action is discounted and

%% Need to clean the above up

A second possibility is that pessimistic evaluation may arise as a result of biased offline replay of events. During periods of rest and sleep, the hippocampus replays recent experiences\citep{whats a good ref?}. A recent theoretical model of replay suggests that these replay events are able to iteratively backpropagate value from a focal point to antecedent states, and that episodes are prioritized for replay according to their likelihood in changing behavior. Biases in replay behavior, or in the mechanisms initiate replay, might result in repeated offline reinforcement of negative episodes thereby exaggerating the value of threat and diffusing its influence across an agent's internal cognitive map. Few studies have investigated replay behavior in aversive environments (e.g. \citep{wu2017}), so this hypothesis remains speculative at this point in time.

%% Want to add anything about transition probability bias?

It is important to note that the present model may not describe all anxiety disorders with equal accuracy. Indeed, our model of pessimistic sequential evaluation is, by definition, a model of prospective cognition. As such, the present results are more likely to accurately describe the anxiety disorders which primarily involve aberrations in future-oriented cognitive processes, such as generalized and social anxiety disorders. Naturally, the present model can neither account for the compulsive behaviors of obsessive-compulsive disorder nor the memory disturbances of post-traumatic stress disorder (though recent clinical studies suggest that diminished perceived control is a vulnerability factor common to all anxiety disorders, including OCD and PTSD\citep{gallagher2014a, gallagher2014b}). Importantly, other psychiatric disorders that involve future-oriented misbeliefs, worry, and avoidance behaviors (e.g. eating disorders\citep{konstantellou2011}) may similarly be well-described by the current account. As such, the present results are transdiagnostic and not limited to one particular diagnosis.

Finally, we are not the first to propose a formal theory of control in psychiatry using MDPs. Huys \& Dayan \cite{HuysDayan2009} also provided a computational account of learned helplessness through simple models of one-step action-outcome contingencies. Our accounts differ particularly in our exclusive focus on control in the sequential setting, which Huys \& Dayan did not address. Indeed, we propose that ultimately key to anxiety is precisely the way in which evaluation in sequential tasks is necessarily reliant on expectations about future choice and events.
